---
title: "Legendary Pokemon Analysis"
author: "Vinay Joshy"
date: "`r Sys.Date()`"
output: html_document
---


<style type="text/css">
  body{
  font-size: 12pt;
  font-family: 'Monaco', 'Courier New', monospace;
}
</style>


```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
```

# Introduction

This report provides an analysis of the Pokemon dataset using R and various statistical and machine learning techniques. The analysis includes data exploration, visualization, and the development of predictive models to determine if a Pokemon is legendary. The data was obtained from [here](https://www.kaggle.com/datasets/rounakbanik/pokemon/)

Let's first load necessary packages:

```{r packages}
library(tidyverse)    # data manipulation and visualization
library(rpart)        # decision trees
library(rpart.plot)   # decision tree plotting
library(randomForest) # random forest packages
library(ROCR)         # classifcations + AUC 

```

# Data Exploration

We can start by importing the dataset, please change path as needed.

```{r}
# Import dataset
pokedex <- read_csv("data/pokemon.csv", 
                    col_types = cols(name = col_factor(), 
                                     type1 = col_factor(),
                                     is_legendary = col_factor()))
```

## Dataset Overview

```{r}
# Display the first six rows
head(pokedex)

# Examine the structure
glimpse(pokedex)

```

### Legendary Pokemon Proportion

```{r}

# Prepare the data
legendary_pokemon <- pokedex %>% 
  count(is_legendary) %>% 
  mutate(prop = n / nrow(pokedex))

# Print the data frame
head(legendary_pokemon)

```

## Visualisations

Creating visualisations is a useful way of identifiying trends in our data.

### Pokemon Distribution by Height and Weight

```{r}

# Prepare the plot
legend_by_heightweight_plot <- pokedex %>% 
  ggplot(aes(x = height_m, y = weight_kg)) +
  geom_point(aes(color = is_legendary), size = 2, alpha = 0.5) +
  geom_text(aes(label = ifelse(height_m > 7.5|weight_kg > 600, 
                            as.character(name), '')), vjust = 0, hjust = 0) +
  geom_smooth(method = "lm", se = FALSE, col = "black", linetype = "dashed") +
  expand_limits(x = 16) +
  labs(title = "Legendary Pokemon by height and weight",
       x = "Height (m)",
       y = "Weight (kg)") +
  guides(color = guide_legend(title = "Pokemon status")) +
  scale_color_manual(labels = c("Non-Legendary", "Legendary"),
                     values = c("#F8730D", "#00eFC4")) +
  theme_bw()

# Print the plot
legend_by_heightweight_plot

```

### Proportion of Legendary Pokemon by Type

```{r}
# Prepare the data
legend_by_type <- pokedex %>% 
    group_by(type1) %>% 
    mutate(is_legendary = as.numeric(is_legendary) - 1) %>% 
    summarise(prop_legendary = mean(is_legendary)) %>% 
    ungroup() %>% 
    mutate(type1 = fct_reorder(type1, prop_legendary))

# Prepare the plot
legend_by_type_plot <- legend_by_type %>% 
    ggplot(aes(x = type1, y = prop_legendary, fill = prop_legendary)) + 
    geom_col() +
    labs(title = "Legendary Pokemon by type") +
    coord_flip() +
    guides(fill = "none") + theme_bw()

# Print the plot
legend_by_type_plot
```

### Pokemon Fight Statistics

```{r}
# Prepare the data
legend_by_stats <- pokedex  %>% 
  select(is_legendary, attack, sp_attack, defense, sp_defense, hp, speed)  %>% 
  gather(key = "fght_stats", value = "value", -is_legendary) 

# Prepare the plot
legend_by_stats_plot <- legend_by_stats %>% 
 ggplot(aes(x = is_legendary, y = value, fill = is_legendary)) +
 geom_boxplot(varwidth = TRUE) +
 facet_wrap(~fght_stats) +
 labs(title = "Pokemon fight statistics",
        x = "Legendary status") +
 guides(fill = "none") + theme_bw()

# Print the plot
legend_by_stats_plot

```

# Machine Learning Analysis

## Data  Splitting

```{r}

# Set seed for reproducibility
set.seed(1234)

# Save number of rows in dataset
n <- nrow(pokedex)

# Generate 60% sample of rows
sample_rows <- sample(n, 0.6 * n)

# Create training set
pokedex_train <- pokedex  %>% 
  filter(row_number() %in% sample_rows) 

# Create test set
pokedex_test <- pokedex  %>% 
  filter(!row_number() %in% sample_rows) %>%
  na.omit()

```


## Decision Tree Model

```{r}
# Load packages and set seed
set.seed(1234)

# Fit decision tree
model_tree <- rpart(is_legendary ~ attack + defense + height_m + 
                    hp + sp_attack + sp_defense + speed + type1 + weight_kg,
                       data = pokedex_train,
                       method = "class",
                       na.action = na.omit)

# Plot decision tree
rpart.plot(model_tree)

```

## Random Forest Model

```{r}
# Load package and set seed
set.seed(1234)

# Fit random forest
model_forest <- randomForest(is_legendary ~ attack + defense + height_m + 
                         hp + sp_attack + sp_defense + speed + type1 + weight_kg,
                         data = pokedex_train,
                         importance = TRUE,
                         na.action = na.omit)

# Print model output
model_forest


```


## Model Evaluation - ROC Curves

```{r}
# Create prediction and performance objects for the decision tree
probs_tree <- predict(model_tree, pokedex_test, type = "prob")
pred_tree <- prediction(probs_tree[,2], pokedex_test$is_legendary)
perf_tree <- performance(pred_tree, "tpr", "fpr")

# Create prediction and performance objects for the random forest
probs_forest <- predict(model_forest, pokedex_test, type = "prob")
pred_forest <- prediction(probs_forest[,2], pokedex_test$is_legendary)
perf_forest <- performance(pred_forest, "tpr", "fpr")

# Plot the ROC curves
plot(perf_tree, col = "red", main = "ROC curves")
plot(perf_forest, add = TRUE, col = "blue")
legend(x = "bottomright",  legend = c("Decision Tree", "Random Forest"), fill = c("red", "blue"))

```

## Random Forest Variable Importance

```{r}
# Print variable importance measures
importance_forest <- importance(model_forest)
importance_forest

# Create a dotchart of variable importance
varImpPlot_forest <- varImpPlot(model_forest)
varImpPlot_forest

```

# Conclusion

In conclusion, this analysis provides insights into the distribution of Pokemon by various attributes and explores the predictive capabilities of decision tree and random forest models. The ROC curves and variable importance measures offer valuable information about the performance and key factors influencing the models.